---
title: "Blood methylation pipeline using recountmethylation on EPIC array - V1"
author:
-  Liuhan Dai
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: bibliography.bib
csl: cell-numeric.csl
package: recountmethylation
output:
  BiocStyle::html_document:
    code_folding: show
    toc: yes
    tocfloat: yes
  BiocStyle::pdf_document: 
    toc: yes
    toc_depth: 2
vignette: > 
  %\VignetteIndexEntry{Blood_EPIC_v1}
  %\VignetteDepends{RCurl}
  %\usepackage[UTF-8]{inputenc} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

# Introdcution and overview: 

This script is to combine functionalities of `BloodMethylationPipeline_EPIC_v0_a` and `v0_b`. What it does is as follows:

1. load two compiled datasets stored on lab server.
  a. Both datasets are measured on illumina infinium methylation EPIC array
  b. Their url on online server is 
  1) https://recount.bio/data/remethdb_h5se-gm_epic_0-0-2_1589820348/
  2) https://recount.bio/data/remethdb_epic-hm850k_h5se_gm_1669220613_0-0-3/

2. Subset blood and whole blood samples for each of them

3. Subset BCAT1 promoter cg probes for blood and whole blood subgroup

4. Visualization and statistical summary

# Environment setup
```{r setup, echo = FALSE, warning = FALSE}
suppressMessages(library(knitr))
suppressMessages(library(GenomicRanges))
suppressMessages(library(limma))
suppressMessages(library(minfi))
suppressMessages(library(ExperimentHub))
suppressMessages(library(recountmethylation))
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = TRUE, 
  message = TRUE)
```

# Load cached datasets on lab server

```{r}
BE.load <- function(fname,cachepath = "Z:/Liuhan Dai/DNA Methylation Detection/20230408 figure 6d") {
bfc <-BiocFileCache(cache = cachepath,ask = interactive())
if (grepl('C:',bfccache(bfc))) q()

filepath <- file.path(bfccache(bfc), fname)
# For this file, getdb() and get_rmdl() neither recoginize its file name.
# Therefore, decide to cache it manually
if (file.exists(filepath) & !fname %in% bfcinfo(bfc)$rname) {
      bfcadd(bfc,fname,fpath =  filepath, action="asis")
} else {
  print(paste(fname,'doesn\'t exist or is already cached'))
}
# Load data by H5SE method
gm <- HDF5Array::loadHDF5SummarizedExperiment(file.path(bfccache(bfc), fname))
return(gm)
}

# load `remethdb_h5se-gm_epic_0-0-2_1589820348` as gm1 and `remethdb_epic-hm850k_h5se_gm_1669220613_0-0-3` as gm2
gm1 <- BE.load('remethdb_h5se-gm_epic_0-0-2_1589820348')
gm2 <- BE.load('remethdb_epic-hm850k_h5se_gm_1669220613_0-0-3')

md1 <-  colData(gm1) # get sample metadata (DO NOT CONVERT TO DATA FRAME!!!!! using as.data.frame)
md2 <-  colData(gm2) # get sample metadata (DO NOT CONVERT TO DATA FRAME!!!!! using as.data.frame)
```

# Subset blood and whole blood samples for `gm1` and `gm2`

```{r}
# extract GSM ID subset for blood and whole blood in `gm1`

md1_blood <- md1[grepl('blood',md1$tissue),]
unique(md1_blood$tissue)
md1_whole_blood <- md1[grepl('whole_blood',md1$tissue),]
unique(md1_whole_blood$tissue)
f <- function(x,y) nrow(md1_whole_blood[y == x,])
sapply(unique(md1_whole_blood$sex), f, y =  md1_whole_blood$sex); sapply(unique(md1_whole_blood$disease), f, y =  md1_whole_blood$disease)
```

```{r}
# extract GSM ID subset for blood and whole blood in `gm2`

md2_blood <- md2[grepl('blood',md2$tissue),]
unique(md2_blood$tissue)
md2_whole_blood <- md2[grepl('whole_blood',md2$tissue),]
unique(md2_whole_blood$tissue)
f <- function(x,y) nrow(md2_whole_blood[y == x,])
sapply(unique(md2_whole_blood$sex), f, y =  md2_whole_blood$sex); sapply(unique(md2_whole_blood$disease), f, y =  md2_whole_blood$disease)
```


Then, let us subset GenomicMethylSet object for whole_blood subgroup using gsm
IDs we have in metadata.

```{r}
BE.subset <- function(gm, md_whole_blood) {
gm_whole_blood <- gm[, gm$gsm %in% md_whole_blood$gsm] # filt h5se object. Here gsm and other meta data is in "column"
# However, once you extract metadata by colData() -> md or md_whole_blood where its metadata now is stored in row
md_whole_blood <- md_whole_blood[order(match(md_whole_blood$gsm, gm_whole_blood$gsm)),] # Synchronize order of md_whole_blood and gr_whole_blood by gsm sequence

if (identical(gm_whole_blood$gsm, md_whole_blood$gsm)) {
    return(gm_whole_blood)
  } else {
    print('GSM seuqence order doesn\'t match')
  }
}

gm1_whole_blood <- BE.subset(gm1,md1_whole_blood)
gm2_whole_blood <- BE.subset(gm2,md2_whole_blood)
```

Secondly, let us extract DNAm Beta-value fractions for a subset of 3 probes that
are within BCAT1 promoter region.

# Subset BCAT1 promoter cg probes for blood and whole blood subgroup in `gm1` and `gm2`

```{r}

BE.BCAT1 <- function(gm_whole_blood) {
# save bval table
bval <- getBeta(gm_whole_blood)
bval <- t(bval) # get transpose
cg_BCAT1 <- c('cg02765913', 'cg23036244', 'cg10764357')
f <- function (x,y)  y[grepl(x,y)]
cgf_BCAT1 <- sapply(cg_BCAT1, f, y = colnames(bval)) # search for which probe for BCAT1 promoter exists in this  dataset
cgf_BCAT1 <- as.character(cgf_BCAT1) ; cgf_BCAT1 <- cgf_BCAT1[nchar(cgf_BCAT1) == 10]
bval_BCAT1 <- bval[,colnames(bval) %in% cgf_BCAT1]  # extract a subset of 3 cg probes that are within BCAT1 promoter region
rownames(bval_BCAT1) <- gsub("\\..*", "", rownames(bval_BCAT1)) # format rownames
BCAT1 <- as.data.frame(bval_BCAT1);
if (identical(gm_whole_blood$gsm, rownames(BCAT1))) {
    BCAT1 <- bind_cols(BCAT1,as_tibble(colData(gm_whole_blood)))
  }
BCAT1 <- as_tibble(BCAT1)

return(BCAT1)
}

gm1_BCAT1 <- BE.BCAT1(gm1_whole_blood) 
gm2_BCAT1 <- BE.BCAT1(gm2_whole_blood)
```

Finally, let us combine both datasets for whole blood subgroup in a single tibble object

# Combine two datasets in a single tibble object gm_BCAT1
```{r}
if (identical(colnames(gm1_BCAT1),colnames(gm2_BCAT1))) {
  if (length(intersect(gm1_BCAT1$gsm, gm2_BCAT1$gsm)) == 0) {
     gm_BCAT1 <- bind_rows(gm1_BCAT1,gm2_BCAT1)
  } else {
    print('These two datasets have overlapping GSM IDs')
  }
} else {
  print('Datasets have different colnames')
  # only combine common columns of gm1_BCAT1 and gm2_BCAT2 to make them consistent with each other
  print('Only select common coldata')
  com_col <- intersect(colnames(gm1_BCAT1),colnames(gm2_BCAT1))
  gm1_BCAT1<- gm1_BCAT1[,com_col] ; gm2_BCAT1<- gm2_BCAT1[,com_col]
}

if (identical(colnames(gm1_BCAT1),colnames(gm2_BCAT1))) {
  if (length(intersect(gm1_BCAT1$gsm, gm2_BCAT1$gsm)) == 0) {
     gm_BCAT1 <- bind_rows(gm1_BCAT1,gm2_BCAT1)
     gm_BCAT1 <- arrange(gm_BCAT1, 'gsm')
  }
}
```

To get a quantitative assessment, we can also create a summary table for gm_BCAT1

```{r}
cg_BCAT1 <- c('cg02765913', 'cg23036244', 'cg10764357')

BCAT1_Sum <- function(BCAT1, cg) {
  sex_summary <- BCAT1 %>%
  group_by(sex) %>%
  summarize(mean_value = mean(.data[[cg]]),
            median_value = median(.data[[cg]]),
            min = min(.data[[cg]]),
            max = max(.data[[cg]])
            )

disease_summary <-
  BCAT1 %>%
  group_by(disease) %>%
  summarize(mean_value = mean(.data[[cg]]),
            median_value = median(.data[[cg]]),
            min = min(.data[[cg]]),
            max = max(.data[[cg]])
  )
sex_summary
disease_summary

# diseaseSex_summary <-
#   BCAT1 %>%
#   group_by(disease, sex) %>%
#   summarize(mean_value = mean(.data[[cg]]),
#             median_value = median(.data[[cg]]),
#             min = min(.data[[cg]]),
#             max = max(.data[[cg]])
#   )
# 
# diseaseSex_summary
}

BCAT1_Sum(gm_BCAT1,cg_BCAT1[3])

```
