---
title: "recountmethylation User's Guide"
author:
- Liuhan
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: bibliography.bib
csl: cell-numeric.csl
package: recountmethylation
output:
  BiocStyle::html_document:
    code_folding: show
    toc: yes
    tocfloat: yes
  BiocStyle::pdf_document: 
    toc: yes
    toc_depth: 2
vignette:  
  %%\VignetteIndexEntry{recountmethylation User's Guide}
  %\VignetteDepends{RCurl}
  %\usepackage[UTF-8]{inputenc} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
```{r setup, echo = FALSE, warning = FALSE}
suppressMessages(library(knitr))
suppressMessages(library(GenomicRanges))
suppressMessages(library(limma))
suppressMessages(library(minfi))
suppressMessages(library(ExperimentHub))
suppressMessages(library(recountmethylation))
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, 
  message = FALSE)
```

# There are three ways of loading data
# 1. load .h5 file from ExperimentHub - caching database
# Method 1 is not very useful since we have to deal with .h5 file directly instead of SummarizedExperiment object
# 2. load from recountmethylation dataserver - full list ( seems to be more complete) get_servermatrix() , getdb
# Method 2 is caching 
# 3. load from local directory - not using cache (I am not so sure whether it is inferior to the first two methods.)
#HDF5Array::load....
geo.red <- geo.red[order(match(rownames(geo.red), rownames(h5.red))),]
# here is a explanation of what this command is doing.
# We start from geo.red[1], find which h5.red[j] (starting from j = 1) matches geo.red[1], return h5.red[j] as the 1st element in output
# In the end, we 

# In the validation step, they first check whether probe ID matches and then they check whether signal matrix matches or not.
# A summary of data processing from geo is as below:
# 1) load by get_IDAT2gr - recountmethylation - RGChannel type
# 2) normalization by minfi::preprocessnoob - Methyltype
# 3) calculate Beta value using getBeta - matrix type
# 4) 

# The code that load anything from recountmethylation data server is as follows:
```{r, eval = FALSE}
# sm <- as.data.frame(smfilt(get_servermatrix())) # This smfilt() will lose some .h5 dataset which is essential for data processing change it to the next line:

sm <- as.data.frame(get_servermatrix())
if(is(sm, "data.frame")){knitr::kable(sm, align = "c")} # display available database on the server

## load methylset - MethylSet files
gmdn <- "remethdb-h5se_gm_0-0-1_1590090412"

# for methylset in data anlaysies vignette, use getdb_h5se type
gm <- getdb_h5se_gm(
  platform = c("hm450k", "epic"),
  dfp = NULL,
  namematch = gmdn,
  verbose = FALSE
)

## load grset - GenomicRanges files
grdn <- "remethdb-h5se_gr_0-0-1_1590090412"
gr <- getdb_h5se_gr(
  platform = c("hm450k", "epic"),
  dfp = NULL,
  namematch = grdn,
  verbose = FALSE
)

```

If you can not find from recountmethylation data server,
The code that load anything from ExperimentHub data server on Bioconductor is as follows:
```{r, eval = FALSE}
cache.path <- tools::R_user_dir("recountmethylation")
setExperimentHubOption("CACHE", cache.path)
hub <- ExperimentHub::ExperimentHub()                    # connect to the hubs
rmdat <- AnnotationHub::query(hub, "recountmethylation") # query the hubs

# for data anlaysies vignette, search ExperimentHub unique identifier for gmdn and grdn
title <- mcols(rmdat)$title

## load methylset
gmdn_UID <- names(rmdat)[grepl("remethdb-h5se_gm_0-0-1_1590090412",title)]

gmdn_fpath <- rmdat[[gmdn_UID]] # download with default caching
gmdn <- rhdf5::h5ls(gmdn_fpath)         # load the h5 file
gm <- loadHDF5SummarizedExperiment(gmdn)

## load grset
grdn_UID <- names(rmdat)[grepl("remethdb-h5se_gm_0-0-1_1590090412",title)]
grdn_fpath <- rmdat[[grdn_UID]] # download with default caching
grdn <- rhdf5::h5ls(grdn_fpath)         # load the h5 file
gr <- loadHDF5SummarizedExperiment(grdn)
```


Several tips as follows:
1. In principl, it might be a good idea to look at metadata from HDF5-SummarizedExperiment type dataset to separate cancer from normal tissues or pick up a specific tissue like whole blood for analysis. Please refere to data_analyses.Rmd
  a. There are some filters before begining data processing.
  b. We can use the disease column in the metadata for whole blood analysis!!!
